### 2012-08-31 Dail_Mad_ext.tex
formatting

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}

@@

These models allow three sources of zero counts by observers: a
species was at a route but not detected; the route was within the
species' range but there were no birds at that site in that year; and
the route was outside the species' range.  Furthermore, detection,
abundance, and zero-inflation can be modeled separately as functions
of different (or the same) covariates.  For example, detection of
species x at site y in year z might depend on wind speed, abundance on
forest type and weather, and zero-inflation upon elevation and
climate.  This approach combines elements of occupancy modeling
(MacKenzie et al. 2006) and abundance modeling, and could prove useful
for distinguishing species' fundamental and realized niches
(Hutchinson 1957).

The zero inflation factor can also affect
recruitment or population dynamics (see below).
We also considered a zero-inflated negative binomial model of initial
abundance, but this model did not perform well in preliminary tests.
Population growth models
Preliminary DM model runs for several species tended to lead to
estimates of survival that were unrealistically high and recruitment
that were unrealistically low, or the reverse (compared to independent
demographic analyses).  The DM models are able to partition changes in
abundance to survival and recruitment in part by making strong
distributional assumptions (Dail and Madsen 2011).  When those
assumptions are heavily violated the models may proportion population
growth incorrectly into survival and recruitment, even if they
estimates population growth accurately.
Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, equation 4 is
simplified to:
(7)
where r represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
(8)
where K is the stable equilibrium of the population and r is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This model is based
on Ricker's (1954) discrete time version of the logistic population
growth model.  We also implemented Gompertz's (1825) density-dependent
model:
(9)
where the log of Nit-1 is only taken when Nit-1 is greater than 0
(when Nit-1 is 0, the full expression simplifies to 0 anyway).  Here
the interpretations of r and K are similar to in the Ricker model, but
K is constrained to be greater than one.
Because a single Poisson distribution controls the
distribution of Nit in each of these models, the discrete
convolution used by Dail and Madsen (2011) is not required,
speeding up processing time.
Immigration and colonization models
The autoregressive, population growth, Ricker, and Gompertz
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The population growth + immigration
model is:
(10)
or
(11)
where $\iota$ represents the immigration rate (equations 10 and 11 are
equivalent).  This model is close to the constant DM model (equation
4), with exp(r) instead of $\omega$ and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial.  Similarly, the
Ricker + immigration model can be represented as:
(12)
the Gompertz + immigration model as:
(13)
and the autoregressive + immigration model as:
(14)
These immigration models are based on the assumption that numbers
of immigrants are not dependent on local density.  Other versions
of these models with different assumptions are possible. The
Ricker + immigration and Gompertz + immigration models can be
justified by the further assumption that immigrants arrive as
adults immediately before the point-count survey (so that
density-dependent processes do not have a chance to act on the
numbers of immigrants; Otto and Day 2007 p. 161).  Deterministic
versions of these models do reach stable equilibriums, but the
equilibriums are not at K and cannot be solved for analytically
(Otto and Day 2007).  We have implemented all preceding models in
a maximum likelihood framework by extending the unmarked package
(Fiske and Chandler 2011) in R (R Development Core Team 2011).
Random eEffects of oObservers
Differences in observers' ability to see, hear, or identify
birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS (Robbins et
al. 1986, Sauer et al. 1994, Diefenbach et al. 2003, Alldredge
et al. 2007, Campbell and Francis 2011).  Estimating a
separate detection probability for each observer can be
difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).
Current BBS trend estimators deal with this problem by
treating observer identity as a random (as opposed to a fixed)
effect (Link and Sauer 2002, Sauer and Link 2011).  This
allows observer-specific differences in detection probability,
but assumes that observers are selected at random from a pool
of potential observers.  Models that contain both random and
fixed effects are referred to as mixed models.  Often in mixed
models (as in this case), the random effect is modeled not
because it of interest in itself, but to avoid bias in the
estimates of the fixed effects.
To include random observer effects in DM models and these extensions,
equation 5 can be modified to:
(15)
(16)
where nijt is the number of bird recorded at site i by observer j in
year t, pj is observer-specific detection probability, $\mu_p$ is the mean
detection probability (on the logit scale), and $\sigma_p$ is the standard
deviation of the random observer effects (also on the logit scale).
We have implemented random observer effects in the Bayesian framework
using program JAGS (version 3.2.0; Plummer 2003) with the R (R
Development Core Team 2011) package rjags (Plummer 2011) as an
interface.
Dynamic range models
The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth terms.
For example, the recruitment term of the constant DM model (equation
4) can be modified as follows:
(17)
When both initial abundance and dynamics are modeled as zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
modeled as varying over space).  In the latter case, it makes sense
for the model to "remember" which sites are inside the range of the
species between the initial abundance stage and each time step of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, trend, Ricker, or Gompertz models
(without immigration) should have no effect when initial abundance is
also zero-inflated unless $\psi$ is time-varying (because these dynamic
models cannot recover from zero abundance).  We have implemented
zero-inflated dynamics in the Bayesian framework using program JAGS
(version 3.2.0; Plummer 2003)

@@

Results

>>>>

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}

@@

These models allow three sources of zero counts by observers: a
species was at a route but not detected; the route was within the
species' range but there were no birds at that site in that year; and
the route was outside the species' range.  Furthermore, detection,
abundance, and zero-inflation can be modeled separately as functions
of different (or the same) covariates.  For example, detection of
species x at site y in year z might depend on wind speed, abundance on
forest type and weather, and zero-inflation upon elevation and
climate.  This approach combines elements of occupancy modeling
\citep{mackenzie_etal:2006} and abundance modeling.
%, and could prove useful
%for distinguishing species' fundamental and realized niches
%(Hutchinson 1957).

The zero inflation factor can also affect
recruitment or population dynamics (see below).
We also considered a zero-inflated negative binomial model of initial
abundance, but this model did not perform well in preliminary tests.

\subsection{Population growth models}

Preliminary DM model runs for several species tended to lead to
estimates of survival that were unrealistically high and recruitment
that were unrealistically low, or the reverse (compared to independent
demographic analyses).  The DM models are able to partition changes in
abundance to survival and recruitment in part by making strong
distributional assumptions. % (Dail and Madsen 2011).
When those assumptions are heavily violated the models may proportion
population growth incorrectly into survival and recruitment, even if
they estimates population growth accurately.

Although partitioning population growth into survival and recruitment
is informative, it is not needed for all applications.  Furthermore, a
simpler model would have several merits: faster running time, fewer
total model combinations, and possibly more realistic estimates.
Therefore, we developed a version of the DM model that estimates and
models population growth directly.  In this model, Eq.\ref{eq:Nt} is
simplified to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1})
\label{eq:exp}
\end{equation}
where $r$ represents the instantaneous population growth rate.  This,
like the autoregressive version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is the
instantaneous population growth rate at low population densities, and
both parameters are constrained to be positive.  This model is based
on the \citet{ricker:1954} discrete time version of the logistic population
growth model.  We also implemented the \citet{gompertz:1825} density-dependent
model:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1})/\log(K)))
\label{eq:gomp}
\end{equation}
where the log of $N_{it-1}$ is only taken when $N_{it-1}>0$
(when $N_{it-1} \equiv 0$, the full expression simplifies to 0 anyway).  Here
the interpretations of $r$ and $K$ are similar to in the Ricker model, but
$K$ is constrained to be greater than one.

Because a single Poisson distribution controls the
distribution of $N_{it}$ in each of these models, the discrete
convolution used by \citet{dail_madsen:2011} is not required,
speeding up processing time.

\subsection{Immigration models}

The autoregressive, population growth, Ricker, and Gompertz
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The population growth plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1}) + \text{Poisson}(\iota)
  \label{eq:expimm1}
\end{equation}
or equivalently
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate.  This model is close to
the constant DM model (equation 4), with $\exp(r)$ instead of $\omega$
and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial. The Ricker and
Gompertz models can be extended to allow for immigration in the same
way.
\begin{comment}
  Similarly, the Ricker + immigration model can be represented as: the
  Gompertz + immigration model as: (13) and the autoregressive +
  immigration model as: (14)
\end{comment}

% Save for discussion
\begin{comment}
  These immigration models are based on the assumption that number of
  immigrant is not dependent on local density; however, extending
  these models in this way is conceptually straight-forward, a Other
  versions of these models with different assumptions are
  possible. The Ricker + immigration and Gompertz + immigration models
  can be justified by the further assumption that immigrants arrive as
  adults immediately before the point-count survey (so that
  density-dependent processes do not have a chance to act on the
  numbers of immigrants; Otto and Day 2007 p. 161).  Deterministic
  versions of these models do reach stable equilibriums, but the
  equilibriums are not at K and cannot be solved for analytically
  (Otto and Day 2007).
\end{comment}

We have implemented all preceding models in
a maximum likelihood framework by extending the \texttt{unmarked} package
\citep{fiske_chandler:2011} in \textbf{R} \citep{R-2012}.

\subsection{Random effects of observers}

Differences in observers' ability to see, hear, or identify
birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS (Robbins et
al. 1986, Sauer et al. 1994, Diefenbach et al. 2003, Alldredge
et al. 2007, Campbell and Francis 2011).  Estimating a
separate detection probability for each observer can be
difficult and reduces one's ability to estimate the quantities
of interest.  This problem is compounded by the fact that
observers differ greatly in the number of surveys they have
run (so that many observers' separate detection probabilities
could not be accurately estimated).

Current BBS trend estimators deal with this problem by
treating observer identity as a random (as opposed to a fixed)
effect (Link and Sauer 2002, Sauer and Link 2011).  This
allows observer-specific differences in detection probability,
but assumes that observers are selected at random from a pool
of potential observers.  Models that contain both random and
fixed effects are referred to as mixed models.  Often in mixed
models (as in this case), the random effect is modeled not
because it of interest in itself, but to avoid bias in the
estimates of the fixed effects.

To include random observer effects in DM models and these extensions,
equation 5 can be modified to:
(15)
(16)
where nijt is the number of bird recorded at site i by observer j in
year t, pj is observer-specific detection probability, $\mu_p$ is the mean
detection probability (on the logit scale), and $\sigma_p$ is the standard
deviation of the random observer effects (also on the logit scale).
We have implemented random observer effects in the Bayesian framework
using program JAGS (version 3.2.0; Plummer 2003) with the R (R
Development Core Team 2011) package rjags (Plummer 2011) as an
interface.

\subsection{Dynamic range models}

The zero-inflated Poisson distribution can be applied to not only
initial abundance but also to recruitment and population growth terms.
For example, the recruitment term of the constant DM model (equation
4) can be modified as follows:
(17)
When both initial abundance and dynamics are modeled as zero-inflated,
$\psi$ can be modeled as time-varying or not (in either case, $\psi$ can also be
modeled as varying over space).  In the latter case, it makes sense
for the model to "remember" which sites are inside the range of the
species between the initial abundance stage and each time step of the
dynamics stage (see code sample?).  Adding zero-inflation to the
dynamics of the autoregressive, trend, Ricker, or Gompertz models
(without immigration) should have no effect when initial abundance is
also zero-inflated unless $\psi$ is time-varying (because these dynamic
models cannot recover from zero abundance).  We have implemented
zero-inflated dynamics in the Bayesian framework using program JAGS
(version 3.2.0; Plummer 2003)

@@

\section{Results}
